Ornaments as indicators of social changes in northeastern Taiwan before and after European colonial period

### 2019-12-17 analysis/paper/paper.Rmd
rearrange the order of figures and add cross referencing

The European presence in northern Taiwan started with the Spanish who founded Fort San Salvador at Keelung in 1626, and Fort San Domingo in 1629 at Tamsui. They sent missionaries to local indigenous settlements in this region [@Blusse2000, pp. 343] and kept records about their observations of indigenous communities. A Dominican priest in 1632 reported that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers used carnelian beads as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is further indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans in the tribe would use carnelian beads as magical items in ritual healing practices [@Borao2009, pp. 122-151]. Records of an indigenous funeral document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate their family's higher status [@LiandWu2006, pp. 153]. While a full critical analysis of these historical accounts remains to be produced, we take them to minimally indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tributes [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on negotations with indigenous communities to secure alliances in the annual ceremony or during travelling [@Kang2016, ch. 6]. We might predict that the activities of the Dutch feudal system to build and maintain alliances resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

@@

@@

```{r tidy-data, eval = TRUE}
# summarise the ornaments by layers
# L5 and L6 represent pre-European contact period, 
# L4 was the time of European contact, and 
# L2 and L1 was the period of Chinese contact. 
ornaments_period <- 
kwl_upper %>% 
  filter(!is.na(`6-layers`)) %>% 
  mutate(period = case_when(
    `6-layers` %in% 1:2 ~ "Chinese Presence",
    `6-layers` == 4 ~ "European Presence",
    `6-layers` %in% 5:6 ~ "Before European Contact",
     TRUE ~ "other"
  )) %>% 
  filter(period != "other") %>% 
  mutate(period = factor(period,
                       level = c("Before European Contact", 
                                 "European Presence", 
                                 "Chinese Presence"))) %>% 
  mutate(Length = as.numeric(`Length(mm)`),
         Thick = as.numeric(`Thick(mm)`),
         Width = as.numeric(`Width(mm)`),
         Perforation1 = as.numeric(`Perforation1(mm)`)) %>% 
  rowwise() %>% 
  mutate(Perforation_average =
           mean(c(`Perforation1`,`Perforation2(mm)`), 
                na.rm = TRUE))
```

@@

Information about Kiwulan (Figure \@ref(fig:KWL-map)) comes from a rescue archaeology project that during 2001-2004 in advance of a water diversion project and road bridge construction. The excavations used 2 mm and 1.5 mm mesh screens and covered eight open area sections in total of 262 squares (4 m by 4 m) reaching 3,814 m^2^ [@Chen2007]. The nearly 2 m thick archaeological deposits reveal a large amount of artifacts, burials, middens, post-holes, wooden pillars, and stone structures, all of which indicates it was a long-term settlement. Artifact locations were recorded to the 2 x 2 m sub-square they were recovered in; they lack individual point provenance. Based on the continuity of deposition and the frequency of artifacts, the center of the site is the open area consisting of A and D sections, which is also the study area where our samples selected from. In the AD area, post-holes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction (Figure \@ref(fig:KWL-sam-area)).

@@

@@

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "small bead", "small (0.5-1 cm)", Type),
         Type = ifelse(Type == "glass bead", "medium (1-2 cm)", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type),
         Categories = ifelse(Categories == "Agate bead", "Carnelian bead", Categories),
         Categories = ifelse(Categories == "Golden bead", "Gold-foil bead", Categories),
         Categories = ifelse(Categories == "Bell", "Metal bell", Categories))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artifact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)

@@

Ornaments were found in a variety of archaeological contexts including post-holes area, burials, and middens. This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aligned post-holes with _in-situ_ posts (Figure \@ref(fig:KWL-sam-area)). Occupation floors were not identified during excavation. We choose these units because they were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site. There are 35 burials in the sampling area, one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, with the total number of ornaments in burials including 3,173. The high number of ornaments in these burials is due to the presence of bead strands or patterned bands of beads, which sometimes contains thousands of beads in an individual burial [@Chen2007]. We exclude burials from this analysis because most of them date to the pre-European period (n = 26), limiting the usefulness of comparisons between the periods. In addition to `r sam_post_hole`  ornaments from dwelling place contexts and 3,173 from burial contexts, there are 27 ornaments found in midden contexts. We focus on ornaments from the dwelling place contexts (Figure \@ref(fig:plot-all-ornaments-count), Table \@ref(tab:tidy-data-for-table)) because these give us the greatest spatial and temporal representation across the three time periods, and so are most informative on social inequality indicated by uneven distributions of ornaments.

>>>>

The European presence in northern Taiwan started with the Spanish who founded Fort San Salvador at Helping dao, Keelung in 1626, and Fort San Domingo in 1629 at Tamsui (Figure \@ref(fig:KWL-map)). They sent missionaries to local indigenous settlements in this region [@Blusse2000, pp. 343] and kept records about their observations of indigenous communities. A Dominican priest in 1632 reported that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers used carnelian beads as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is further indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans in the tribe would use carnelian beads as magical items in ritual healing practices [@Borao2009, pp. 122-151]. Records of an indigenous funeral document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate their family's higher status [@LiandWu2006, pp. 153]. While a full critical analysis of these historical accounts remains to be produced, we take them to minimally indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tributes [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on negotations with indigenous communities to secure alliances in the annual ceremony or during travelling [@Kang2016, ch. 6]. We might predict that the activities of the Dutch feudal system to build and maintain alliances resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

@@

Information about Kiwulan (Figure \@ref(fig:KWL-map)) comes from a rescue archaeology project that during 2001-2004 in advance of a water diversion project and road bridge construction. The excavations used 2 mm and 1.5 mm mesh screens and covered eight open area sections in total of 262 squares (4 m by 4 m) reaching 3,814 m^2^ [@Chen2007]. The nearly 2 m thick archaeological deposits reveal a large amount of artifacts, burials, middens, post-holes, wooden pillars, and stone structures, all of which indicates it was a long-term settlement. Artifact locations were recorded to the 2 x 2 m sub-square they were recovered in; they lack individual point provenance. Based on the continuity of deposition and the frequency of artifacts, the center of the site is the open area consisting of A and D sections, which is also the study area where our samples selected from (Figure \@ref(fig:KWL-sam-area)). In the AD area, post-holes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction.

@@

```{r tidy-data, eval = TRUE}
# summarise the ornaments by layers
# L5 and L6 represent pre-European contact period, 
# L4 was the time of European contact, and 
# L2 and L1 was the period of Chinese contact. 
ornaments_period <- 
kwl_upper %>% 
  filter(!is.na(`6-layers`)) %>% 
  mutate(period = case_when(
    `6-layers` %in% 1:2 ~ "Chinese Presence",
    `6-layers` == 4 ~ "European Presence",
    `6-layers` %in% 5:6 ~ "Before European Contact",
     TRUE ~ "other"
  )) %>% 
  filter(period != "other") %>% 
  mutate(period = factor(period,
                       level = c("Before European Contact", 
                                 "European Presence", 
                                 "Chinese Presence"))) %>% 
  mutate(Length = as.numeric(`Length(mm)`),
         Thick = as.numeric(`Thick(mm)`),
         Width = as.numeric(`Width(mm)`),
         Perforation1 = as.numeric(`Perforation1(mm)`)) %>% 
  rowwise() %>% 
  mutate(Perforation_average =
           mean(c(`Perforation1`,`Perforation2(mm)`), 
                na.rm = TRUE))

# some basic stat for personal ornament section 
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)
```

@@

@@

Ornaments (Figure \@ref(fig:ornaments-image)) were found in a variety of archaeological contexts including post-holes area, burials, and middens. This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aligned post-holes with _in-situ_ posts (Figure \@ref(fig:KWL-sam-area)). Occupation floors were not identified during excavation. We choose these units because they were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site. There are 35 burials in the sampling area, one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, with the total number of ornaments in burials including 3,173. The high number of ornaments in these burials is due to the presence of bead strands or patterned bands of beads, which sometimes contains thousands of beads in an individual burial [@Chen2007]. We exclude burials from this analysis because most of them date to the pre-European period (n = 26), limiting the usefulness of comparisons between the periods. In addition to `r sam_post_hole` ornaments from dwelling place contexts and 3,173 from burial contexts, there are 27 ornaments found in midden contexts. We focus on ornaments from the dwelling place contexts (Figure \@ref(fig:plot-all-ornaments-count), Table \@ref(tab:tidy-data-for-table)) because these give us the greatest spatial and temporal representation across the three time periods, and so are most informative on social inequality indicated by uneven distributions of ornaments.

@@

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "small bead", "small (0.5-1 cm)", Type),
         Type = ifelse(Type == "glass bead", "medium (1-2 cm)", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type),
         Categories = ifelse(Categories == "Agate bead", "Carnelian bead", Categories),
         Categories = ifelse(Categories == "Golden bead", "Gold-foil bead", Categories),
         Categories = ifelse(Categories == "Bell", "Metal bell", Categories))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artifact counts ")

@@
